Evolutionary Algorithm for the Bonded Molecular Conformation Problem

### 2018-08-31 documentation/paper.tex
Finished paper

\documentclass{article}
\usepackage[style=ieee]{biblatex}
\bibliography{references.bib}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

The model molecular energy minimisation problem involves the geometric
configuration of a chain of atoms in two dimensions, each connected by rigid
bonds of unit length, such that the total potential energy yielded by the sum of
individual Lennard-Jones potentials is globally minimised. The total potential
energy of the cluster is given by
$$V(r)=\sum^{N-1}_{i=1}\sum^{N}_{j=i+1}\left(r^{-12}_{ij}-2r^{-6}_{ij}\right)$$
Where $r$ is the euclidean distance between a pair of atoms and $N$ is the
number of atoms in the chain.

The energy minimisation problem is historically difficult because of the
exponential relationship between the locally stable structures and the number of
atoms, $N$, in the molecule. That is, as the size of the cluster increases, the
number of local minima increases exponentially such that when $N$ approaches
100, there exist more than $10^{140}$ minima \cite{DAVEN1996195}. It is
therefore necessary to employ sophisticated search algorithms in order to make
larger input sizes attainable.

Section~\ref{litreview} will discuss some of the research that has already taken
place in this area. Section~\ref{proposed} will detail the algorithm proposed by
this paper.

@@

As simulated annealing has been found to be successful when combined with
evolutionary algorithms in similar problem domains \cite{PhysRevLett.91.080201},
this paper will explore its use as an intermediate non-gradient method such that
the proposed algorithm has three levels: global meta-heuristic (GA),
intermediate non-gradient optimisation (SA) and local gradient-based
optimisation (BFGS). The performance of this three stage optimisation method
will be compared with a purely evolutionary algorithm without the intermediate
step.

@@

The mutation mechanism is where the intermediate non-gradient optimisation
algorithm will be incorporated. In traditional genetic algorithms, mutation has
a small chance of occurring and is random, resulting in some states which are
less fit in the hopes that this will later lead to a deeper basin in the search
space. By incorporating simulated annealing at the mutation stage it is hoped
that the configuration will be dragged in a downward direction while also
providing enough stochastic properties to allow boundaries between minima to be
traversed, which is the goal of any mutation operator.

@@

This mutation operator is based on one described by Pullan \cite{PULLAN1998331}
in which A random angle index $\alpha_i$ is selected in the molecule and the
distance between this point and each endpoint is compared. The angles between
the selected angle and the closest end forms a new segment of angles. For each
angle in the selected segment, a very small (close to zero) random number is
chosen and the angle is set to that value. This results in one end of the
molecule being `unravelled' into a straight line, which effectively removes any
knots or kinks within that segment.The segment is later `rewound' through
subsequent local optimisation.

@@

A random angle is selected and both the current fitness of the molecule and the
current value of this angle is stored.The selected angle is set to a random
value between $-\pi$ and $\pi$ and the fitness is recomputed. If the new fitness
is worse than the previous fitness, the angle is reverted to its previous state.
This process is then repeated until either the maximum number of iterations is
reached, or an improved configuration is found. This limit is important because
after many generations it becomes increasingly difficult to find a random change
in an angle that does not give lower energy, which would otherwise result in an
infinite loop.

\paragraph{Naive Mutation}

@@

A relatively brief annealing schedule is used due to the fact that during every
iteration an energy computation is required. The goal at this intermediate
search algorithm is not to find the global minimum, but merely to provide a more
localised search for deeper basins in the vicinity of the current configuration;
it is hoped that if the GA search has found a local minimum in close proximity
to the global minimum, the SA stage will take large enough steps in order to
climb over the boundary into the neighbouring basin where it can be finally
optimised by the gradient search in the next stage. Therefore, an annealing
schedule that prioritises speed over optimality will be chosen. The initial
temperature $T$ will be set to 100 and decrement by one every iteration until it
reaches 0.

During annealing, random angles are selected and modified by random amounts. If
the change produces an improved state, it is always accepted. If the change
produces a worsened state, it is accepted with a probability of $1 - e^{-\Delta
E / Tk}$ such that both the current `temperature' as well as the degree to which
the change will worsen the configuration, determine whether the change is to be
accepted. $\Delta E$ represents the change in energy that will occur if the new
configuration is accepted, $T$ represents the current temperature, and $k$ is
used to scale probability curve. A value of $k=0.03$ was chosen as this provides
a reasonable probability curve for energy deltas of 1 and below.

@@

The following GA parameters are used:
\begin{description}
	\item[Population Size] 12
	\item[Maximum Generations] 100
	\item[Mutation Rate] Deterministic and based on conditions as described in
	section~\ref{mutationrate}
\end{description}
The GA works by continually sorting a large pool (of size $n_{pop} \times 2$) of
configurations in order to rank them in order of fitness and subsequently select
states for crossover and mutation operators. The population is initialised with
random molecules and sorted and then the procedure outlined in REF are carried
out until either a known global optimum is found or the maximum number of
generations is reached.

%%%% GA Pseudocode here

The final stage of each GA iteration, after candidate configurations have been
found by the GA and its immediate vicinity has been further explored through
mutation, is to refine the configuration using gradient-based local search. This
stage uses the gradient of the objective function in order to guarantee that the
bottom of the current basin will be reached. The quasi-Newton method
Broyden–Fletcher–Goldfarb–Shanno (BFGS) will be applied at this stage resulting
in a fast descent towards the current local minimum at each stage. The
coordinates of this locally optimised configuration are also be passed back to
the GA so that improvements are retained in future iterations. This is known as
a `Lamarkian'process and has previously been found to be beneficial when used
with this problem \cite{doi:10.1002/qua.560440214}.

\section{Results}

@@

\section{Conclusion}

@@

%\subsection{Implementation}

% \begin{algorithm} \caption{Proposed High-Level Algorithm} \label{algorithm}
%   \centering \begin{algorithmic} \Procedure{optimise}{} % \State $nn \gets new
%   Network$ %   \For{$i \gets 0; i <= number\_of\_epochs; i++$} %
%   \For{$\textit{each data partition of minibatch size}$} %           \State
%   $\textit{Do forward pass}$ %           \State $\textit{Backpropagate the
%   resulting error}$ %           \State $\textit{Update the weights
%   proportionally to their contribution to the error}$ %       \EndFor %
%   \State $\textit{Shuffle and repartition the data}$ %   \EndFor \State $P
%   \gets \textit{Initial Population Size}$ \State $population \gets initialise$
%   \EndProcedure \end{algorithmic}

% \end{algorithm}

>>>>

\documentclass{article}
\usepackage[style=ieee]{biblatex}
\bibliography{references.bib}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{listings}
\usepackage{amsmath}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}

@@

The model molecular energy minimisation problem involves the geometric
configuration of a chain of atoms in two dimensions, each connected by rigid
bonds of unit length, such that the total potential energy yielded by the sum of
individual Lennard-Jones potentials is globally minimised. The total potential
energy of the system is given by
$$V(r)=\sum^{N-1}_{i=1}\sum^{N}_{j=i+1}\left(r^{-12}_{ij}-2r^{-6}_{ij}\right)$$
Where $r$ is the euclidean distance between a pair of atoms and $N$ is the
number of atoms in the chain.

The bound energy minimisation problem is historically difficult because of the
exponential relationship between the locally stable structures and the number of
atoms, $N$, in the molecule. That is, as the size of the cluster increases, the
number of local minima increases exponentially such that when $N$ approaches
100, there exist more than $10^{140}$ minima \cite{DAVEN1996195}. It is
therefore necessary to employ global search algorithms in order to make
larger input sizes attainable.

The primary difficulty encountered with the bonded variation of the problem (as
opposed to the unbound `Lennard-Jones cluster' problem), is that when atomic
bonds overlap, a large amount of energy is required to pull an atom through the
knot \cite{doi:10.1002/qua.560440214}. This is particularly problematic when
local optimisation algorithms are applied, as they will not move towards the
higher energy states required for knots to be removed. Due to this limitation of
local search methods, a global algorithm is required in order to facilitate the
traversal between local minima in the global search space by effectively finding
unknotted configurations.

Section~\ref{litreview} will discuss relevant research that has already taken
place in this area. Section~\ref{proposed} will detail the algorithm proposed by
this paper.

@@

As simulated annealing has been found to be successful when combined with
evolutionary algorithms in similar problem domains \cite{PhysRevLett.91.080201},
this paper will explore its use as an intermediate non-gradient method such that
the proposed algorithm has three levels: global meta-heuristic (GA),
intermediate non-gradient optimisation (SA) applied here in the form of a
genetic
mutation function, and local gradient-based optimisation (BFGS). The performance
of this three stage optimisation method will be compared with a purely
evolutionary algorithm without the intermediate step.

@@

The mutation mechanism is where the intermediate non-gradient optimisation
algorithm is incorporated. In traditional genetic algorithms, mutation has
a small chance of occurring and is random, resulting in some states which are
less fit in the hopes that this will later lead to a deeper basin in the search
space. By incorporating simulated annealing at the mutation stage it is expected
that the configuration will be dragged in a downward direction while also
providing enough stochastic properties to allow boundaries between minima to be
traversed, which is the goal of any mutation operator.

@@

This mutation operator is based on one described by Pullan \cite{PULLAN1998331}
in which a random angle index $\alpha_i$ is selected in the molecule and the
distance between this point and each endpoint is compared. The angles between
the selected angle and the closest end forms a new segment of angles. For each
angle in the selected segment, a very small (close to zero) random number is
chosen and the angle is set to that value. This results in one end of the
molecule being `unravelled' into a straight line, which effectively removes any
knots or kinks within that segment.The segment is later `rewound' through
subsequent local optimisation.

@@

A random angle is selected and both the current fitness of the molecule and the
current value of this angle is stored.The selected angle is set to a random
value between $-\pi$ and $\pi$ and the fitness is recomputed. If the new fitness
is worse than the previous fitness, the angle is reverted to its previous state.
This process is then repeated until either the maximum number of iterations is
reached, or an improved configuration is found. This limit is important because
after many generations it becomes increasingly difficult to find a random change
in an angle that does not increase the energy, which without the iteration limit
would result in an infinite loop.

\paragraph{Naive Stochastic Mutation}

@@

A relatively brief annealing schedule is used due to the fact that during every
iteration an energy computation is required. The goal of this intermediate
search algorithm is not to find the global minimum, but merely to provide a more
localised search for deeper basins in the vicinity of the current configuration;
it is hoped that if the GA search has found a local minimum in close proximity
to the global minimum, the SA stage will take large enough steps in order to
climb over the boundary into the neighbouring basin where it can be finally
optimised by the gradient search in the next stage. Therefore, an annealing
schedule that prioritises speed over optimality will be chosen. The initial
temperature $T$ will be set to 100 and decrement by one every iteration until it
reaches 0.

During annealing, random angles are selected and modified by random amounts. If
the change produces an improved state, it is always accepted. If the change
produces a worsened state, it is accepted with a probability of $1 - e^{-\Delta
E / Tk}$ such that both the current `temperature', as well as the degree to which
the change will worsen the configuration, determine whether the change is to be
accepted. $\Delta E$ represents the change in energy that will occur if the new
configuration is accepted, $T$ represents the current temperature, and $k$ is
used to scale probability curve. A value of $k=0.03$ was chosen as this provides
a reasonable probability curve for energy deltas of 1 and below.

@@

The following GA parameters are used:
\begin{description}
	\item[Population Size] 12
	\item[Maximum Generations] 100
	\item[Mutation Rate] Deterministic and based on conditions as described in
	section~\ref{mutationrate}
\end{description}
This study compares the performance of the GA using an annealed mutation
operator with that using only a purely stochastic operator. Both tests also use
the guided mutation and the `unravel' mutation as possible operators to be
selected at random.

The GA works by continually sorting a large pool (of size $n_{pop} \times 2$) of
configurations in order to rank them in order of fitness and subsequently select
states for crossover and mutation operators. The population is initialised with
random molecules and sorted and then the procedure outlined in
algorithm~\ref{pseudo} are carried out until either a known global optimum is
found or the maximum number of generations is reached.

\begin{algorithm}[H] \label{pseudo}
	initialise population\;
	\For{$\leftarrow 0$ \KwTo $max\_gen$} {
		\For{$c\leftarrow n\_pop \KwTo n\_pop * 2$} {
			$p1\leftarrow Population[rand \% pop\_size / 2]$\;
			$p2\leftarrow Population[rand \% pop\_size / 2]$\;
			$Population[c] \leftarrow Crossover(p1, p2)$\;
		}
		\ForEach{s in Population} {
			$s.fitness \leftarrow calculateFitness()$\;
		}
		$quicksort(Population)$\;
		\For{$i \leftarrow 1 \KwTo n\_pop * 2$} {
			\If {$Population[i] - Population[i-1] < 1.0 \textit{OR} Population
			[i].fitness > 0$} {
				$randomMutation(Population[i])$\;
			}
		}
		\ForEach{s in Population} {
			$BFGS(s)$\;
		}
		$quicksort(Population)$
	}
\end{algorithm}
The final stage of each GA iteration, after candidate configurations have been
found by the GA and their immediate vicinities have been further explored
through mutation, is to refine the configuration using gradient-based local
search. This stage uses the gradient of the objective function in order to
guarantee that the bottom of the current basin will be reached. The quasi-Newton
method Broyden–Fletcher–Goldfarb–Shanno (BFGS) will be applied at this stage
resulting in a fast descent towards the current local minimum at each stage. The
coordinates of this locally optimised configuration are also be passed back to
the GA so that improvements are retained in future iterations. This is known as
a `Lamarkian'process and has previously been found to be beneficial when used
with this problem \cite{doi:10.1002/qua.560440214}.

\section{Results \& Conclusion}

@@

The genetic algorithm was tested in two configurations for comparison:
\begin{itemize}
	\item With the `unravel' mutation operator, the guided mutation operator and
	the purely stochastic operator
	\item With the `unravel' and guided mutation operators but replacing the
	stochastic operator with the annealed operator.
\end{itemize}
It was expected that the annealed mutation would yield more optimal results
for higher $N$ values by providing a means of traversing barriers between local
minima and steeper basins within their vicinity. However, as seen in figure~
\ref{fig:nvse}, while optimal results were found equally well by both
configurations for around $n \leq 30$, at larger $n$ values the configuration
with the annealed mutation operator performed less consistently and found
global minima less often than the version with the stochastic mutation operator
and no SA.

The SA mutation operator was successful in its goal of extending the search to
configurations
surrounding those found by the GA, which was apparent in the increased variation
within populations compared to the relatively homogeneous populations of the
stochastic version. However, this increased variation may
have been detrimental to the overall performance of the algorithm. Given the
relatively high rate of mutation and the nature of the simulated annealing
method, multiple worsened states are accepted for each mutation stage as opposed
to the stochastic version which can only accept one worsened state per
mutation. In other words, the simulated annealing approach is too aggressive
for use as a mutation operator because early in the annealing process it accepts
potentially many worsened states, throwing away important information found by
the GA.

It is thought that perhaps a less aggressive acceptance function or applying SA
at a different stage in the algorithm (not as a mutation operator) may negate
these effects, but it is apparent that the genetic algorithm approach provides
sufficient global exploration, making the SA stage redundant.

@@
